T gates, temporary Toffolis, and halving the cost of quantum computation

### 2017-08-23 paper.tex
Fleshing out the logical-and section

\section{AND-ing and Erasing}
\label{sec:temporary-and}

When the target of a Toffoli gate is a qubit in the $|0\rangle$ state, and the controls are in a computation basis state, the target qubit ends up storing the logical-and of the two controls.
We continue to say that the target stores the logical-and of the two controls even when they are under superposition, with the understanding that we mean ``within each computation basis state making up the superposition of the overall system".

In the textbook construction of the Toffoli gate, the three T gates on the control wires are implementing a controlled-S.
A controlled-S phases states by $i$ where both the control and the target are ON.
If the target of the Toffoli gate starts in the $|0\rangle$ state, then the target ends up ON when both controls are satisfied.
Therefore the target is ON in exactly the state where a phase factor of $i$ should be applied, and the controlled-S on the controls can be replaced by an unconditional S gate on the target after the Toffoli is finished.

A minor optimization which we apply is to note that the first two operations on the target are transitioning from the $|0\rangle$ state to the $|A\rangle$ state.
Instead of consuming an $|A\rangle$ state to perform the $T$ gate (and potentially performing a fixup S gate), we simply feed the $|A\rangle$ state we would have consumed directly into the circuit.

This construction defines our logical AND operation, which we draw as a wire emerging from two controls as shown in the left hand side of \autoref{fig:logical-and} (which also shows our construction for the operation).

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=1.5em @C=0.7em {
      &x &&\ctrl{1} &\qw & x  && &&          &&\qw      &\qw  &\ctrl{2} &\qw &\qw      &\qw  &\qw &\qw &\qw &\\
      &y &&\ctrl{1} &\qw & y  &&=&&          &&\ctrl{1} &\qw  &\qw      &\qw &\ctrl{1} &\qw  &\qw &\qw &\qw &\\
      &  &&         &\qw & xy && &&|A\rangle &&\targ    &\qTi &\targ    &\qT &\targ    &\qTi &\qH &\qS &\qw &\\
    }
  }
  \caption{
	Control merging with T-cost of 4 (three T gates, one $|A\rangle$ state).
	The $|A\rangle$ state is fed directly into the circuit, instead of using an $H$ and $T$ gate on a $|0\rangle$ state (which would consume an $|A\rangle$ state anyways).
	The input $|A\rangle$ state can be recovered by using the un-merging construction in \autoref{fig:unmerge-recover-a}, reducing the net T cost of this circuit to 3.
	However, the un-merging construction in \autoref{fig:unmerge-free} is a better choice, despite not recovering the $|A\rangle$ state, because the total T-cost of merge+unmerge using \autoref{fig:unmerge-free} is 4+0=4 instead of 4+3-1=6.
  }
  \label{fig:logical-and}
\end{figure}

@@

To erase the ancilla produced by the logical AND, we use a technique we refer to as ``erasure".
We note that a Toffoli gate would obviously clear the value.
Then, since the target qubit can be discarded after clearing it, we are free to introduce a Hadamard gate and a measurement before discarding it.
We then hop the Hadamard over the Toffoli, transforming it into a CCZ operation.
The controls and targets of a CCZ are equivalent, so we rewrite the circuit into a CZC.
We then hop the Measurement over the CZC, turning a quantum control into a classical control.
These circuit moves turned the Toffoli gate into a CZ gate that we apply or not based on the outcome of the measreument
We ``Clifford-ized" the circuit by adding a Hadamard gate and a measurement.

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=1.5em @C=0.7em {
      &x  &&\ctrl{1} &\qw &x && &&\qw &\qw    &\ctrl{1} &\qw \\
      &y  &&\ctrl{1} &\qw &y &&=&&\qw &\qw    &\gate{Z} &\qw \\
      &xy &&\qw      &    &  && &&\qH &\meter &\cw \cwx \bullet &    \\
    }
  }
  \caption{
	Control un-merging with T-cost of 0.
  }
  \label{fig:erase-and}
\end{figure}

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=1.5em @C=0.7em {
      &\ctrl{1} &\qw && &&\ctrl{1} &\qw      &\ctrl{1} &\qw && &&          &&\qw &\qw &\qw      &\qw  &\ctrl{2} &\qw &\qw      &\qw  &\ctrl{2} &\qw &\qw  &\qw      &\qw &\qw    &\ctrl{1}         &\qw \\
      &\ctrl{2} &\qw && &&\ctrl{1} &\qw      &\ctrl{1} &\qw && &&          &&\qw &\qw &\ctrl{1} &\qw  &\qw      &\qw &\ctrl{1} &\qw  &\qw      &\qw &\qw  &\qw      &\qw &\qw    &\gate{Z}         &\qw \\
      &         &    &&=&&         &\ctrl{1} &\qw      &    &&=&&|0\rangle &&\qH &\qT &\targ    &\qTi &\targ    &\qT &\targ    &\qTi &\targ    &\qH &\qSi &\ctrl{1} &\qH &\meter &\cw \cwx \bullet &    \\
      &\targ    &\qw && &&\qw      &\targ    &\qw      &\qw && &&          &&\qw &\qw &\qw      &\qw  &\qw      &\qw &\qw      &\qw  &\qw      &\qw &\qw  &\targ    &\qw &\qw    &\qw              &\qw \\
    }
  }
  \caption{
	Performing a Toffoli gate by computing, using, and erasing a logical-AND.
	Has a T-cost of 4, and is exactly equivalent to the construction from \cite{Jones2013}.
  }
  \label{fig:merge-use-erase-toffoli}
\end{figure}

>>>>

\section{Computing and Erasing the Logical-AND}
\label{sec:temporary-and}

A Toffoli gate whose target is in the $|0\rangle$ state is the quantum equivalent of the logical-AND operation.
For each computational basis state, with the controls storing the bits $x$ and $y$, the target ends up storing $x \land y$.

In the previous section, we showed how to compute and uncompute the logical-AND of two controls by investing an $|A\rangle$ state and performing three $T$ gates.
This construction be improved in several ways.

First, the computation is introducing phase errors (as shown in \autoref{fig:bad-phase-toffoli}).
Normally, correcting these phase errors would require controlled-Z on the target (applied before the Toffoli) and a controlled-S between the two controls.
We know our target is $|0\rangle$ beforehand, so the controlled-Z has no effect and can be ignored.
The effect of the controlled-S gate is to apply a phase factor of $i$ to the amplitudes of computational basis states where both controls on.
In other words, the controlled-S phases the logical-and of the two controls and we just happen to have a qubit storing that value.
We can correct computation's phase error by applying an uncontrolled S gate to the ancilla storing the logical-and.

Second, the construction is unnecessarily deep.
Every $T$ gate is associated with an S gate that will happen based on the result of a measurement.
It takes time for the measurement result to become available, and it is beneficial to be waiting for several results in parallel instead of one after another.
We can reduce our measurement depth to 1 using the same technique as in \cite{Jones2013} (except we only need to perform three T gates instead of four, thanks to having the $|A\rangle$ state as an input).

Third, taking another hint from \cite{Jones2013}, we can perform the uncomputation by using an erasure technique.
To erase the ancilla produced by the logical AND, we use a technique we refer to as ``erasure".
We note that a Toffoli gate would obviously clear the value.
Then, since the target qubit can be discarded after clearing it, we are free to introduce a Hadamard gate and a measurement before discarding it.
We then hop the Hadamard over the Toffoli, transforming it into a CCZ operation.
The controls and targets of a CCZ are equivalent, so we rewrite the circuit into a CZC.
We then hop the Measurement over the CZC, turning a quantum control into a classical control.
These circuit moves turned the Toffoli gate into a CZ gate that we apply or not based on the outcome of the measreument
We ``Clifford-ized" the circuit by adding a Hadamard gate and a measurement.

With these three improvements applied, we have the three key ingredients of our temporary Toffoli with a T-cost of 4: computing the logical-and, using the logical-and, and uncomputing the logical-and.

We draw the computation of the logical and as a wire emerging vertically from its two inputs (drawn as controls) and heading rightward, as shown in \autoref{fig:compute-logical-and}.
We draw the uncomputation analogously, with a wire storing the logical-and coming in from the left then merging vertically into its two inputs (drawn as controls), as shown in \autoref{fig:erase-logical-and}.

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=1.5em @C=0.7em {
      &x &&\ctrl{1} &\qw & x  && &&          &&\targ    &\ctrl{2}&\qw     &\qTi &\qw     &\ctrl{2}&\targ     &\qw &\qw &\qw &\\
      &y &&\ctrl{1} &\qw & y  &&=&&          &&\targ    &\qw     &\ctrl{1}&\qTi &\ctrl{1}&\qw     &\targ     &\qw &\qw &\qw &\\
      &  &&         &\qw & xy && &&|A\rangle &&\ctrl{-2}&\targ   &\targ   &\qT  &\targ   &\targ   &\ctrl{-2} &\qH &\qS &\qw &\\
    }
  }
  \caption{
	Computing the logical-and of two qubits, with a T-cost of 4 and a measure-react depth of 1.
  }
  \label{fig:compute-logical-and}
\end{figure}

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=1.5em @C=0.7em {
      &x  &&\ctrl{1} &\qw &x && &&\qw &\qw    &\ctrl{1} &\qw \\
      &y  &&\ctrl{1} &\qw &y &&=&&\qw &\qw    &\gate{Z} &\qw \\
      &xy &&\qw      &    &  && &&\qH &\meter &\cw \cwx \bullet &    \\
    }
  }
  \caption{
	Uncomputing the logical-and of two qubits, with a T-cost of 0 and a measure-react depth of 1.
  }
  \label{fig:erase-logical-and}
\end{figure}

During the time that the logical-and ancilla wire exists, it can be used to control operations onto other qubits.
In effect, these operations are doubly-controlled by the inputs that produced the ancilla.
For example, we can perform a Toffoli by computing the logical-and of the two controls, applying a CNOT from the logical-and to the actual target, then uncomputing the logical-and.
This construction, shown in \autoref{fig:merge-use-erase-toffoli}, is equivalent to the one in \cite{Jones2013} (but broken into pieces that can be used independently).

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=1.5em @C=0.7em {
      &\ctrl{1} &\qw && &&\ctrl{1} &\qw      &\ctrl{1} &\qw && &&          &&\targ    &\ctrl{2}&\qw     &\qTi &\qw     &\ctrl{2}&\targ     &\qw &\qw &\qw      &\qw &\qw    &\ctrl{1}         &\qw \\
      &\ctrl{2} &\qw && &&\ctrl{1} &\qw      &\ctrl{1} &\qw && &&          &&\targ    &\qw     &\ctrl{1}&\qTi &\ctrl{1}&\qw     &\targ     &\qw &\qw &\qw      &\qw &\qw    &\gate{Z}         &\qw \\
      &         &    &&=&&         &\ctrl{1} &\qw      &    &&=&&|A\rangle &&\ctrl{-2}&\targ   &\targ   &\qT  &\targ   &\targ   &\ctrl{-2} &\qH &\qS &\ctrl{1} &\qH &\meter &\cw \cwx \bullet &    \\
      &\targ    &\qw && &&\qw      &\targ    &\qw      &\qw && &&          &&\qw      &\qw     &\qw     &\qw  &\qw     &\qw     &\qw       &\qw &\qw &\targ    &\qw &\qw    &\qw              &\qw \\
    }
  }
  \caption{
	Performing a Toffoli gate by computing the logical-and of its controls, using the result, then erasing it.
	Has a T-cost of 4 and a measure-react depth of 2.
	Equivalent to the construction from \cite{Jones2013} shown in \autoref{fig:jones-toffoli}.
  }
  \label{fig:merge-use-erase-toffoli}
\end{figure}

@@

@@

@@
